A Survey of Real-Time Strategy Game AI\\ Research and Competition in StarCraft

### 2013-07-29 survey.tex
corrections sections I and II

\section{Introduction}\label{sec:intro}
\IEEEPARstart{T}{he} field of real-time strategy (RTS) game AI has advanced significantly since Michael Buro's call for research in this area \cite{Buro03rts}. Specially, competitions like the ``ORTS RTS Game AI Competition'' (held from 2006 to 2009), and the ``AIIDE StarCraft AI Competition'' (held since 2010) have motivated the exploration of many AI approaches in the context of RTS AI. We will list and classify these approaches, explain their 
strengths and weaknesses and conclude on what is left to achieve human-level 
RTS AI.

Complex dynamic environments, where it is not possible to have neither perfect nor complete information about the current state or about the dynamics of the environment, pose significant challenges for artificial intelligence. Road  traffic, finance, or
weather  forecasts are  examples of such large, complex, real-life dynamic
environments. RTS games can be seen as a simplification of one such real-life
environment, with simpler dynamics in a  finite and smaller
world, although still complex enough  to study some of the key interesting
problems like decision making  under
uncertainty  or  real-time adversarial planning. Finding efficient  techniques for
tackling  these problems  on  RTS  games can  thus  benefit other AI disciplines and application
domains, and also have concrete  and direct applications in  the ever growing
industry of  video games.%, generating worldwide two  and a  half times more revenue than the Film industry in 2011.
% This statement is not fair, the reference I found it only compare all the game industry revenue (including hardware sales) vs theatrical film industry revenue (i.e. not DVD sales, ...)

This paper aims to provide a one-stop guide on what is the state of
the art in RTS AI, with a particular emphasis on the work done in StarCraft. 
It is organized as follows: Section~\ref{sec:rts}
introduces RTS games, in particular the game StarCraft,
and  their main  AI challenges.  Section~\ref{sec:review}
reviews  the existing  work on tackling these challenges in RTS games.
Section~\ref{sec:bot} analyzes several current state of the art RTS game playing agents (called {\em bots}), selected form the participants to annual
StarCraft    AI    competitions.
Section~\ref{sec:competition}  presents results of the recent annual competitions
held at the AIIDE  and CIG conferences and a StarCraft bot game ladder\footnote{An extended tournament, which can potentially go on indefinitely.}.  Section~\ref{sec:questions}
compiles open questions in RTS game AI. Finally, the paper concludes on discussions and perspectives.

@@

\begin{itemize}
	\item Terrans, provide units that are versatile and flexible giving a balanced option between Protoss and Zergs.
	\item Protoss units have lengthy and expensive manufacturing process, but they are strong and resistant. These conditions make players follow a strategy of quality over quantity.
	\item Zergs, the insectoid race, units are cheap and weak. They can be produced fast, encouraging players to overwhelm their opponents with sheer numbers.
\end{itemize}

@@

A typical StarCraft map is defined as a rectangular grid, where the $width \times height$ of the map is measured in the number of $32 \times 32$ squares of pixels, also known as build tiles. Although the
resolution of walkable areas is in squares of $8 \times 8$ pixels, also known as walk tiles. The typical dimensions for maps range from $64 \times 64$ to $256 \times 256$ build tiles. Each player  can
control up  to 200 units (plus an unlimited number of buildings).  Moreover, each
different race  contains between 30  to 35 different types  of units
and  buildings,  most of them with  a significant  number  of  special
abilities. All these factors  together make StarCraft a significant
challenge,   in   which   humans    are   still   much   better   than
computers. For instance, in the game ladder iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked by their current point totals ($E$ being the lowest possible rank, and $A^+$ and $Olympic$ being the second highest and highest ranks, respectively), the best StarCraft AI bots are ranked between $D$ and $D^+$, where average  amateur players are  ranked between $C^+$ and  $B$. For comparison, StarCraft professional  players are usually ranked between
$A^-$ and $A^+$.

From a theoretical point of view, the state space of a StarCraft for a given map is enormous. For example, consider a $128 \times 128$ map. At any given moment there might be between 50 to 400 units in the map, each of which might have a complex internal state (remaining energy and hit-points, action being executed, etc.). This quickly leads to an immense number of possible states (way beyond the size of smaller games, such as Chess or Go). For example, just considering the location of each unit (with $128 \times 128$ possible positions per unit), and 400 units, gives us an initial number of $16384^{400} \approx 10^{1685}$. If we add the other factors playing a role in the game, we obtain even larger numbers.

Another way  to measure the complexity of the game is  by looking at the branching factor, $b$, and the depth of the game, $d$, as proposed in~\cite{Gaby}, with a total game complexity of $b^d$. In Chess, $b \approx 35$ and $d \approx 80$. In more complex games, like Go, $b \approx 30 - 300$, and $d \approx 150 - 200$. In order to determine the branching factor in StarCraft when an AI plays it, we must have in mind, that the AI can issue actions simultaneously to as many units in the game as desired. Thus, considering that, in a typical game, a player controls between 50 to 200 units, the branching factor would be between $u^{50}$ and $u^{200}$, where $u$ is the average number of actions each unit can execute. 
{\color{blue}
Estimating the value of $u$ is not easy, since the number of actions a unit can execute is highly dependent on the context. Let us make the following assumptions: 1) at most 16 enemy units will be in range of a friendly unit (larger values are possible, but are unlikely), 2) when an AI plays StarCraft, it only makes sense to consider movement in the 8 cardinal directions per unit (instead of assuming that the player can issue a ``move'' command to anywhere in the map at any point in time), 3) for ``build'' actions, we consider that SCVs only build in their current location (otherwise, if they need to move, we consider that as first issuing a ``move'' action, and then a ``build''), and 4) let's consider only the {\em Terran} race. With those assumptions, units in StarCraft can execute between 1 (units like ``Supply Depots'', whose only action is to be ``idle'') to 43 actions (Terran ``Ghosts''), with typical values around 20 to 30. Now, if we have in mind that actions have cool-down times, and thus not all units can execute all of the actions at every frame, we can take a conservative estimation of about $10$ possible actions per unit per game frame. This results in a conservative estimate for the branching factor between $b \in [10^{50},10^{200}]$, only considering units (ignoring the actions buildings can execute). 
}
%Assuming a conservative value for $u$ of about $30$ (moving to nearby positions, attacking units in range, using special abilities, etc.), we obtain a branching factor of $b$ between  $30^{50}$ and $30^{200}$.
Now, to compute $d$, we simply consider the fact that typical games last for about 25 minutes, which results in $d \approx 36000$ (25 minutes $\times$ 60 seconds $\times$ 24 iterations per second).

@@

\subsubsection{Spatial and Temporal Reasoning}
% TODO just a paragraph on exposing why there are temporal and spatial
% reasoning in RTS AI.
Spatial  reasoning   is  related  to   each  aspect  of   terrain
exploitation.  Spatial reasoning is involved in tasks such as building placing, or base expansion. In the former, the player needs to
carefully  consider building  position  into its  own  bases to  both
protect them  by creating  a wall against  invasions and to  avoid bad
configurations  where large  units could  be stuck. In base expansion, the player has to
choose good available locations to  build a new base, regarding its
own  position  and opponent's  bases.  Finally,  spatial reasoning  is
key to tactical reasoning: players need to decide where to place units
for battle,  favoring, for  instance, engagements when  the opponent's
units are into a bottleneck.

% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
% [3], and 53.125% according  to BWAPI website. Who's right? Should we
% add this in the paper?]
% [Alberto: I guess Weber's number comes from StarCraft walkthroughs like this: http://www.supercheats.com/pc/walkthroughs/starcraftbroodwar-walkthrough01.txt, there is also a chance to hit in even grounds (like 98%). But since we cannot see the original code all of these numbers are approximations. In any case we can keep explaining the concept without numbers ;)]
Another example of spatial reasoning in StarCraft is that it is always an advantage to  have its units  on high ground  while enemy is on low ground, since units  on low ground have no vision  on high ground, and firing from low ground have a chance to  hit units on high ground.

Analogously, temporal reasoning is key in tactical or strategic reasoning. For example, timing attacks and retreats to gain advantage. At a higher strategic level, players need to reason about when to perform long-term impact economic actions such as upgrades, building construction, strategy switching, etc. all taking into account that the effects of these actions are not immediate, but longer term.

\subsubsection{Domain Knowledge Exploitation}
In traditional board games such as Chess, researchers have exploited the large amounts of existing domain knowledge to create good evaluation functions to be used by alpha-beta search algorithms, extensive opening books, or end-game tables. In the case of RTS games, it is still unclear the ways in which the significantly large amount of domain knowledge (in the forms or strategy guides, replays, etc.) can be exploited by bots. Most work in this area has focused on two main directions: on the one hand, researchers are finding ways in which to hard-code existing strategies into bots, so that bots only need to decide which strategies to deploy, instead of having to solve the complete problem of deciding which actions to execute by each individual unit at each time step. One the other hand, large datasets of replays have been created \cite{WeberCig09,synnaeve2012dataset}, from where strategies, trends or plans have been tried to learn. However, StarCraft games are quite complex, and how to automatically learn from such datasets is still an open problem.

\subsubsection{Task Decomposition}
For all the previous reasons, most existing approaches to play games like StarCraft work by decomposing the problem of playing a RTS game into a collection of smaller problems, to be solved independently. Specifically, a common subdivision is:
\begin{itemize}
\item {\em Strategy}: corresponds to the high-level decision making process.  This is
  the  highest level of  abstraction for  the game  comprehension. Finding   an   efficient   strategy   or
  counter-strategy  against a  given opponent is key in RTS games.  It concerns  the whole
  set of  units and buildings  a player owns. %If  we may allow  ourselves a  military analogy, strategy is orders given by generals at the HQ. 
\item  {\em Tactics}: are  the implementation  of the  current  strategy. It
  implies  army and  building positioning,  movements, timing,  and so
  on.  Tactics concerns a  group of  units. %They  are orders  given by officers on the battlefield.
\item  {\em Reactive  control}: is  the  implementation  of  tactics.  This
  consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
  techniques (also knows as ``kiting'')  during battle. Reactive control focuses
  on a specific unit. %It is soldiers applying officers' orders. 
%\end{itemize}
%
%\noindent
%Both     strategy      and     tactics     are      usually     called
%\textit{macro-management},  whereas reactive  control  is also  called
%\textit{micro-management}. In addition of the three tasks shown above,
%one could also consider these two tasks that an AI has to deal with:
%\begin{itemize}
\item {\em Terrain analysis}: consists  in the analysis of regions composing
  the  map:  choke-points,  minerals  and  gas emplacements,  low  and  high
  walkable grounds, islands, etc.
\item {\em Intelligence gathering}:  corresponds to information collected about
  the  opponent. Because of  the fog-of-war,  players must
  regularly send scouts to localize and spy enemy bases.%, and must see or  infer what buildings and technology the enemy has developed in order  to adapt its strategy.
\end{itemize}

In comparison, when humans play StarCraft, they typically divide their decision making in a very different way. The StarCraft community typically talks about two tasks:
\begin{itemize}
  \item {\em Micro}: is the ability to control your units individually {\color{blue} (roughly corresponding to {\em Reactive Control} above, and part of {\em Tactics})}. A good \emph{micro} player usually keeps their units alive during more time.
  \item {\em Macro}: is the ability to produce units and to expand at the appropriate times to keep your production of units flowing {\color{blue} (roughly corresponding to everything but {\em Reactive Control} and part of {\em Tactics} above)}. A good \emph{macro} player usually has the larger army.
\end{itemize}

>>>>

\section{Introduction}\label{sec:intro}
\IEEEPARstart{T}{he} field of real-time strategy (RTS) game AI has advanced significantly since Michael Buro's call for research in this area \cite{Buro03rts}. Specially, competitions like the ``ORTS RTS Game AI Competition'' (held from 2006 to 2009), the ``AIIDE StarCraft AI Competition'' (held since 2010), and the
``CIG StarCraft RTS AI Competition'' (held since 2011) have motivated the exploration of many AI approaches in the context of RTS AI. We will list and classify these approaches, explain their 
strengths and weaknesses and conclude on what is left to achieve human-level 
RTS AI.

Complex dynamic environments, where neither perfect nor complete information about the current state or about the dynamics of the environment are available, pose significant challenges for artificial intelligence. Road  traffic, finance, or
weather  forecasts are  examples of such large, complex, real-life dynamic
environments. RTS games can be seen as a simplification of one such real-life
environment, with simpler dynamics in a  finite and smaller
world, although still complex enough  to study some of the key interesting
problems like decision making  under
uncertainty  or  real-time adversarial planning. Finding efficient  techniques for
tackling  these problems  on  RTS  games can  thus  benefit other AI disciplines and application
domains, and also have concrete  and direct applications in  the ever growing
industry of  video games.%, generating worldwide two  and a  half times more revenue than the Film industry in 2011.
% This statement is not fair, the reference I found it only compare all the game industry revenue (including hardware sales) vs theatrical film industry revenue (i.e. not DVD sales, ...)

This paper aims to provide a one-stop guide on what is the state of
the art in RTS AI, with a particular emphasis on the work done in StarCraft. 
It is organized as follows: Section~\ref{sec:rts}
introduces RTS games, in particular the game StarCraft,
and  their main  AI challenges.  Section~\ref{sec:review}
reviews  the existing  work on tackling these challenges in RTS games.
Section~\ref{sec:bot} analyzes several current state of the art RTS game playing agents (called {\em bots}), selected from the participants to annual
StarCraft    AI    competitions.
Section~\ref{sec:competition}  presents results of the recent annual competitions
held at the AIIDE  and CIG conferences and a StarCraft bot game ladder\footnote{An extended tournament, which can potentially go on indefinitely.}.  Section~\ref{sec:questions}
compiles open questions in RTS game AI. Finally, the paper concludes on discussions and perspectives.

@@

\begin{itemize}
	\item Terrans provide units that are versatile and flexible giving a balanced option between Protoss and Zergs.
	\item Protoss units have lengthy and expensive manufacturing processes, but they are strong and resistant. These conditions make players follow a strategy of quality over quantity.
	\item Zergs, the insectoid race, units are cheap and weak. They can be produced fast, encouraging players to overwhelm their opponents with sheer numbers.
\end{itemize}

@@

A typical StarCraft map is defined as a rectangular grid, where the $width \times height$ of the map is measured in the number of $32 \times 32$ squares of pixels, also known as build tiles. However, the
resolution of walkable areas is in squares of $8 \times 8$ pixels, also known as walk tiles. The typical dimensions for maps range from $64 \times 64$ to $256 \times 256$ build tiles. Each player  can
control up  to 200 units (plus an unlimited number of buildings).  Moreover, each
different race  contains between 30  to 35 different types  of units
and  buildings,  most of them with  a significant  number  of  special
abilities. All these factors  together make StarCraft a significant
challenge,   in   which   humans    are   still   much   better   than
computers. For instance, in the game ladder iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked by their current point totals ($E$ being the lowest possible rank, and $A^+$ and $Olympic$ being the second highest and highest ranks, respectively), the best StarCraft AI bots are ranked between $D$ and $D^+$, where average  amateur players are  ranked between $C^+$ and  $B$. For comparison, StarCraft professional  players are usually ranked between
$A^-$ and $A^+$.

From a theoretical point of view, the state space of a StarCraft game for a given map is enormous. For example, consider a $128 \times 128$ map. At any given moment there might be between 50 to 400 units in the map, each of which might have a complex internal state (remaining energy and hit-points, action being executed, etc.). This quickly leads to an immense number of possible states (way beyond the size of smaller games, such as Chess or Go). For example, just considering the location of each unit (with $128 \times 128$ possible positions per unit), and 400 units, gives us an initial number of $16384^{400} \approx 10^{1685}$. If we add the other factors playing a role in the game, we obtain even larger numbers.

Another way  to measure the complexity of the game is  by looking at the branching factor, $b$, and the depth of the game, $d$, as proposed in~\cite{Gaby}, with a total game complexity of $b^d$. In Chess, $b \approx 35$ and $d \approx 80$. In more complex games, like Go, $b \approx 30$ to $300$, and $d \approx 150$ to $200$. In order to determine the branching factor in StarCraft when an AI plays it, we must have in mind, that the AI can issue actions simultaneously to as many units in the game as desired. Thus, considering that, in a typical game, a player controls between 50 to 200 units, the branching factor would be between $u^{50}$ and $u^{200}$, where $u$ is the average number of actions each unit can execute. 
{\color{blue}
Estimating the value of $u$ is not easy, since the number of actions a unit can execute is highly dependent on the context. Let us make the following assumptions: 1) at most 16 enemy units will be in range of a friendly unit (larger values are possible, but unlikely), 2) when an AI plays StarCraft, it only makes sense to consider movement in the 8 cardinal directions per unit (instead of assuming that the player can issue a ``move'' command to anywhere in the map at any point in time), 3) for ``build'' actions, we consider that SCVs (terran worker units) only build in their current location (otherwise, if they need to move, we consider that as first issuing a ``move'' action, and then a ``build''), and 4) let's consider only the {\em Terran} race. With those assumptions, units in StarCraft can execute between 1 (units like ``Supply Depots'', whose only action is to be ``idle'') to 43 actions (Terran ``Ghosts''), with typical values around 20 to 30. Now, if we have in mind that actions have cool-down times, and thus not all units can execute all of the actions at every frame, we can take a conservative estimation of about $10$ possible actions per unit per game frame. This results in a conservative estimate for the branching factor between $b \in [10^{50},10^{200}]$, only considering units (ignoring the actions buildings can execute). 
}
%Assuming a conservative value for $u$ of about $30$ (moving to nearby positions, attacking units in range, using special abilities, etc.), we obtain a branching factor of $b$ between  $30^{50}$ and $30^{200}$.
Now, to compute $d$, we simply consider the fact that typical games last for about 25 minutes, which results in $d \approx 36000$ (25 minutes $\times$ 60 seconds $\times$ 24 iterations per second).

@@

\subsubsection{Spatial and Temporal Reasoning}
% TODO just a paragraph on exposing why there are temporal and spatial
% reasoning in RTS AI.
Spatial  reasoning   is  related  to   each  aspect  of   terrain
exploitation.  It is involved in tasks such as building placement or base expansion. In the former, the player needs to
carefully  consider building  positioning  into its  own  bases to  both
protect them  by creating  a wall against  invasions and to  avoid bad
configurations  where large  units could  be stuck. In base expansion, the player has to
choose good available locations to  build a new base, regarding its
own  position  and opponent's  bases.  Finally,  spatial reasoning  is
key to tactical reasoning: players need to decide where to place units
for battle,  favoring, for  instance, engagements when  the opponent's
units are lead into a bottleneck.

% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
% [3], and 53.125% according  to BWAPI website. Who's right? Should we
% add this in the paper?]
% [Alberto: I guess Weber's number comes from StarCraft walkthroughs like this: http://www.supercheats.com/pc/walkthroughs/starcraftbroodwar-walkthrough01.txt, there is also a chance to hit in even grounds (like 98%). But since we cannot see the original code all of these numbers are approximations. In any case we can keep explaining the concept without numbers ;)]
Another example of spatial reasoning in StarCraft is that it is always an advantage to  have own units  on high ground  while the enemy is on low ground, since units  on low ground have no vision  onto the  high ground. %, and firing from low ground have a chance to  hit units on high ground. 
%MP: took out last half sentence as it seems to be rather confusing.

Analogously, temporal reasoning is key in tactical or strategic reasoning. For example, timing attacks and retreats to gain an advantage. At a higher strategic level, players need to reason about when to perform long-term impact economic actions such as upgrades, building construction, strategy switching, etc. all taking into account that the effects of these actions are not immediate, but longer term.

\subsubsection{Domain Knowledge Exploitation}
In traditional board games such as Chess, researchers have exploited the large amounts of existing domain knowledge to create good evaluation functions to be used by alpha-beta search algorithms, extensive opening books, or end-game tables. In the case of RTS games, it is still unclear how the significantly large amount of domain knowledge (in the forms or strategy guides, replays, etc.) can be exploited by bots. Most work in this area has focused on two main directions: on the one hand, researchers are finding ways in which to hard-code existing strategies into bots, so that bots only need to decide which strategies to deploy, instead of having to solve the complete problem of deciding which actions to execute by each individual unit at each time step. One the other hand, large datasets of replays have been created \cite{WeberCig09,synnaeve2012dataset}, from where strategies, trends or plans have been tried to learn. However, StarCraft games are quite complex, and how to automatically learn from such datasets is still an open problem.

\subsubsection{Task Decomposition}
For all the previous reasons, most existing approaches to play games as StarCraft work by decomposing the problem of playing an RTS game into a collection of smaller problems, to be solved independently. Specifically, a common subdivision is:
\begin{itemize}
\item {\em Strategy}: corresponds to the high-level decision making process.  This is
  the  highest level of  abstraction for  the game  comprehension. Finding   an   efficient   strategy   or
  counter-strategy  against a  given opponent is key in RTS games.  It concerns  the whole
  set of  units and buildings  a player owns. %If  we may allow  ourselves a  military analogy, strategy is orders given by generals at the HQ. 
\item  {\em Tactics}: are  the implementation  of the  current  strategy. It
  implies  army and  building positioning,  movements, timing,  and so
  on.  Tactics concerns a  group of  units. %They  are orders  given by officers on the battlefield.
\item  {\em Reactive  control}: is  the  implementation  of  tactics.  This
  consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
  techniques (also knows as ``kiting'')  during battle. Reactive control focuses
  on a specific unit. %It is soldiers applying officers' orders. 
%\end{itemize}
%
%\noindent
%Both     strategy      and     tactics     are      usually     called
%\textit{macro-management},  whereas reactive  control  is also  called
%\textit{micro-management}. In addition of the three tasks shown above,
%one could also consider these two tasks that an AI has to deal with:
%\begin{itemize}
\item {\em Terrain analysis}: consists  in the analysis of regions composing
  the  map:  choke-points,  minerals  and  gas emplacements,  low  and  high
  walkable grounds, islands, etc.
\item {\em Intelligence gathering}:  corresponds to information collected about
  the  opponent. Because of  the fog-of-war,  players must
  regularly send scouts to localize and spy enemy bases.%, and must see or  infer what buildings and technology the enemy has developed in order  to adapt its strategy.
\end{itemize}

In comparison, when humans play StarCraft, they typically divide their decision making in a very different way. The StarCraft community typically talks about two tasks:
\begin{itemize}
  \item {\em Micro}: is the ability to control units individually {\color{blue} (roughly corresponding to {\em Reactive Control} above, and part of {\em Tactics})}. A good \emph{micro} player usually keeps their units alive over a longer period of time.
  \item {\em Macro}: is the ability to produce units and to expand at the appropriate times to keep your production of units flowing {\color{blue} (roughly corresponding to everything but {\em Reactive Control} and part of {\em Tactics} above)}. A good \emph{macro} player usually has the larger army.
\end{itemize}
